---
title: "Longitudinal data analysis"
author: "Nnamdi Joseph Asouzu"
output:
  html_document:
    df_print: paged
---
**LONGITUDINAL DATA MODELLING AND ANALYSIS**<br>

**Case study** <br>

The data come from the study reported by Potthoff and Roy in Biometrika (1964).
In the study the distance from the center of the pituitary to the maxillary fissure was recorded
at ages 8, 10, 12, and 14, for 11 girls and 16 boys. <br>
Scientific question: Is dental growth related to gender?<br>

```{r}
library(lme4)
library(lattice)
library(arm)
library(car)
library(pbkrtest)
```


```{r}
df <- read.table(file="C:/Users/Nnamdi/Desktop/Bioinformatics/Statistical_methods_for_bioinformatics/part_I/Projects/Project4/growth.txt", header=T)
attach(df)
df
```
<br>
**Exploratory data analysis**<br>

Spaghetti plot<br>
```{r}
n=length(unique(IDNR))
interaction.plot(AGE,IDNR,MEASURE, xlab="Age in years", ylab="Distance from the pituitary to the maxillary fissure",legend=F, col = "purple")
```
<br>
There is a general increase in dental growth with increasing age. Also, there is variation in dental growth among individuals within each age group regardless the increasing age. <br>

<br>
Boxplots <br>
```{r}
boxplot(MEASURE~AGE,xlab="Age (in years)",ylab="Distance from the pituitary to the maxillary fissure", col = "red")
```
<br>
The boxplot show overall increase in distance from the center of the pituitary to the maxillary fissure as the individuals get older. <br>
<br>
Boxplots per gender <br>
```{r}
boxplot(MEASURE[SEX==1]~AGE[SEX==1],main="Male gender",main="Male gender",xlab="Age (in years)",ylab="Distance from the pituitary to the maxillary fissure", col = "red")
```
<br>
Overall, there is increase in dental growth with increase in age in males <br>

```{r}
boxplot(MEASURE[SEX==2]~AGE[SEX==2],main="Female gender", main="Female gender",xlab="Age (in years)",ylab="Distance from the pituitary to the maxillary fissure", col = "pink")
```
<br>
Overall, there is increase in dental growth with increase in age in females <br>

<br>
```{r}
hist(df$MEASURE, col="sky blue")
```


<br>
**Model fitting** <br>

```{r}
df.lm <- lmer(MEASURE ~ SEX*AGE + (1 + AGE|IDNR), df, REML = FALSE)
summary(df.lm)
```

```{r}
anova(df.lm)
```
<br>
Estimating fixed effect via bootstrap <br>
```{r}
fixed.boot <- bootMer(df.lm, fixef, use.u=TRUE, nsim=250)
summary(fixed.boot)
```
<br>

Calculating confidence interval for the fixed effects <br>
```{r}
confint(df.lm,method="boot",boot.type ="perc",oldNames = FALSE,nsim=500)
```
The estimates of the fixed effects fall within the confidence interval. <br>

<br>
Get the KR-approximated degrees of freedom <br>
```{r}
df.KR <- get_Lb_ddf(df.lm, fixef(df.lm))
early.lmer1.coef=coef(summary(df.lm))
early.lmer1.p.KR <- cbind(early.lmer1.coef,2 * (1 - pt(abs(early.lmer1.coef[,3]), df.KR)))
early.lmer1.p.KR
```
Age and gender significantly affects the distance from the center of the pituitary to the maxillary fissure. <br>

<br>
Likelihoood ratio tests <br>
```{r}
df.lm.null<-lmer(MEASURE ~ AGE + (1 + AGE|IDNR), df, REML = FALSE)
df.lm.treat<-lmer(MEASURE ~ SEX*AGE + (1 + AGE|IDNR), df, REML = FALSE)
anova(df.lm.null,df.lm.treat)
```
The model with the treatment is significant hence gender affects dental growth. <br>

<br>
The mixed model is given as <br>
**MEASURE ~ AGE + SEX + SEX:AGE + (1 + AGE | IDNR) + e**. <br>










