---
title: "Model fitting and selection using linear regression"
author: "Nnamdi Joseph Asouzu"
date: "8/22/2021"
output:
  html_document:
    df_print: paged
---

**CASE STUDY** <br>

Low birth weight data: The goal of this study was to identify risk factors associated with
giving birth to a low birth weight baby (weighing less than 2500 grams). Data were collected
on 189 women, 59 of which had low birth weight babies and 130 of which had normal birth
weight babies. Four variables which were thought to be of importance were age, weight of
the subject at her last menstrual period, race, and the number of physician visits during the
first trimester of pregnancy.<br>

**ID&emsp;Meaning** <br>
LOW&emsp;Low BirthWeight (0 = BirthWeight >= 2500g, 1 = Birth Weight < 2500g)<br>
AGE&emsp;Age of the Mother in Years <br>
LWT&emsp;Weight in Pounds at the Last Menstrual Period <br>
RACE&emsp;Race (1 = White, 2 = Black, 3 = Other) <br>
SMOKE&emsp;Smoking Status During Pregnancy (1 = Yes, 0 = No) <br>
PTL&emsp;History of Premature Labor (0 = None 1 = One, etc.) <br>
HT&emsp;History of Hypertension (1 = Yes, 0 = No) <br>
UI&emsp;Presence of Uterine Irritability (1 = Yes, 0 = No) <br>
FTV&emsp;Number of Physician Visits During the First Trimester (0 = None, 1 = One, 2 = Two, etc.)<br>
BWT&emsp;Birth Weight in Grams <br>

*Use the continuous variable BWT as the response variable.* <br>

**Analysis 1:** <br>
* Considering the four most clinically important variables previously discussed, i.e., age,
weight of the subject at her last menstrual period, race, and the number of physician
visits during the first trimester of pregnancy, carry out a model building exercise to
determine the best model to describe the data. Based your model building exercise on
the AIC and AIC weights. Discuss your results carefully. 
<br>
* Use backward elimination procedure to decide which predictor variables can be dropped
from the regression model. Control the type I error at alpha = 0.10 at each stage. Which
variables are retained? How does this compare to the results obtained with the AIC? <br>

**Analysis 2:** <br>
* Consider now the 8 covariates included in the data set (all except Birth Weight in
Grams). Repeat the previous model building exercise using the AIC and the AIC
weights. Compare the results of both analysis and make a careful discussion.
* Use backward elimination procedure to decide which predictor variables can be dropped
from the regression model. Control the type I error at alpha = 0.10 at each stage. Which
variables are retained? How does this compare to the results obtained with the AIC? <br>


```{r}
suppressWarnings(suppressMessages(library("pastecs")))
suppressWarnings(suppressMessages(library("tidyverse")))
suppressWarnings(suppressMessages(library("PerformanceAnalytics")))
suppressWarnings(suppressMessages(library("AICcmodavg")))
```

**Import dataset** <br>
```{r}
df <- read.table(file="C:/Users/Nnamdi/Desktop/Bioinformatics/Statistical_methods_for_bioinformatics/part_I/Projects/Project2/lowbwt.dat", header=T)
head(df)
```
<br>
**ANALYSIS 1** <br>

**Descriptive statistics** <br>
```{r}
desc.df <- stat.desc(df[,c("AGE","LWT","RACE","SMOKE","PTL","HT","UI","FTV")], basic =T, desc=T)
desc.df
df$RACE <- as.factor(df$RACE)
df$FTV <- as.factor(df$FTV)
df$SMOKE <- as.factor(df$SMOKE)
df$PTL <- as.factor(df$PTL)
df$HT <- as.factor(df$HT)
df$UI <- as.factor(df$UI)
df$LOW <- as.factor(df$LOW)
```

**Exploratory data analysis**
```{r}
chart.Correlation(df[,c(3,4,11)], histogram=TRUE, pch=19)
```
<br>
The birth weight of a new born baby increases when the age of the mother is 30 and above. This observation is supported by the medical condition called Macrosomia. Likewise, the birth weight of the baby increases with an increase in the weight of the mother before her last menstrual period. <br>

```{r}
ggplot(df,aes(FTV,BWT,color = FTV))+geom_boxplot() + theme_classic()

```
<br>
This plot shows that mothers had at least 6 prenatal visits to the doctor will likely not have babies with low birth weight.<br>

```{r}
ggplot(df,aes(RACE,BWT,color = RACE))+geom_boxplot() + theme_classic()
```
<br>
The plot shows that white women have babies with higher birth weight compared to African women or women from other ethnicity<br>


**Fitting Models** <br>
```{r}
mod <- lm(BWT ~ AGE+LWT+RACE+FTV, data=df)
summary (mod)
```
<br>

**Forward Selection using AICc score**
```{r}
mod.list <- list()
mod.list[[1]] <- lm(BWT ~ AGE+LWT+RACE+FTV, data=df)
mod.list[[2]]<- lm(BWT ~ AGE*FTV+AGE*LWT+AGE*RACE+LWT*RACE+LWT*FTV+RACE*FTV, data = df)
mod.list[[3]] <- lm(BWT ~ AGE+LWT+RACE, data=df)
mod.list[[4]] <- lm(BWT ~ AGE+FTV+RACE, data=df)
mod.list[[5]] <- lm(BWT ~ FTV+LWT+RACE, data=df)
mod.list[[6]] <- lm(BWT ~ AGE+LWT*RACE, data=df)
mod.list[[7]] <- lm(BWT ~ AGE+FTV*RACE, data=df)

mod.aictab=aictab(cand.set = mod.list) 
mod.aictab
```
<br>
The model fitted by forward selection technique using the lowest AICc score is **BWT ~ AGE + LWT + RACE**. <br>


<br>
**Backward selection using cut-off p-value < 0.10** <br>

```{r}
df.mod <- lm(BWT~AGE+LWT+RACE+FTV, data = df)
summary(df.mod)

```
<br>
Remove the variable with p-value above the cut off score <br>
```{r}
df.mod2 <- update(df.mod, . ~ . - FTV, data = df)
summary(df.mod2)
```
<br>
Age has a p -value greater than 0.1 but we will include it in the model because it is known that maternal age affects birth weight of babies ( a condition known as Macrosomia) <br>

Thus, the model fitted by backward selection technique using control p-value less than 0.10 is **BWT ~ AGE + LWT + RACE**. This model is same as the model fitted using the forward selection.  <br> 


**ANALYSIS 2** <br>
<br>

**Exploratory data analysis**
```{r}
chart.Correlation(df[,-c(1,2,5:10)], histogram=TRUE, pch=19)
```
<br>
The birth weight of a new born baby increases when the age of the mother is 30 and above. This observation is supported by the medical condition called Macrosomia. Likewise, the birth weight of the baby increases with an increase in the weight of the mother before her last menstrual period.
<br>

Boxplots<br>
```{r}
ggplot(df,aes(RACE,BWT,color = RACE))+geom_boxplot() + theme_classic()
```
<br>
The plot shows that white women have babies with higher birth weight compared to African women or women from other ethnicity
<br>
```{r}
ggplot(df,aes(SMOKE,BWT,color = SMOKE))+geom_boxplot() + theme_classic()
```
<br>
The barplot shows that women who smoked cigarettes during pregnancy had babies with lower birth weight compared to non-smokers pregnant women.
<br>
```{r}
ggplot(df,aes(PTL,BWT,color = PTL))+geom_boxplot() + theme_classic()
```
<br>
The barplot shows that women with the highest history(3 times) of premature labor had babies with higher weight compared to other women with lesser number of premature labor.
<br>
```{r}
ggplot(df,aes(HT,BWT,color = HT))+geom_boxplot() + theme_classic()
```
<br>
The plot shows that women with hypertension had babies with lower birth weight compared to women without hypertension.
<br>
```{r}
ggplot(df,aes(UI,BWT,color = UI))+geom_boxplot() + theme_classic()
```
<br>
The plot shows that women with urinary tract infections had babies with lower birth weight compared to women without uterine irritability.
<br>
```{r}
ggplot(df,aes(FTV,BWT,color = FTV))+geom_boxplot() + theme_classic()
```
<br>
This plot shows that mothers had at least 6 prenatal visits to the doctor will likely have babies with higher birth weight.<br>

**Fitting Models** <br>

**Model building using forward selection technique and AICc score**
```{r}
mod.list2 <- list()
mod.list2[[1]] <- lm(BWT ~ AGE+LWT+RACE+SMOKE+PTL+HT+UI+FTV, data=df)
mod.list2[[2]]<- lm(BWT ~ LWT+RACE+SMOKE+PTL+HT+UI, data = df)
mod.list2[[3]] <- lm(BWT ~ AGE+LWT+RACE+SMOKE+PTL+HT+UI, data = df)
mod.list2[[4]] <- lm(BWT ~ LWT*RACE+LWT*SMOKE+LWT*PTL+LWT*HT+LWT*UI, data=df)
mod.list2[[5]] <- lm(BWT ~ AGE+RACE*HT+RACE*UI+LWT*RACE+SMOKE*PTL+SMOKE*HT+SMOKE*RACE, data=df)
mod.list2[[6]] <- lm(BWT ~ LWT+RACE+SMOKE+PTL+HT+UI+RACE*UI, data=df)

mod.aictab2=aictab(cand.set = mod.list2) 
mod.aictab2
```

<br>
The model fitted by forward selection technique using AICc score is <br> **BWT ~ AGE+LWT+RACE+SMOKE+PTL+HT+UI**. <br>
Although this model did not have the lowest AICc score, it is logical to include the mother's age as one of the predictors of the birth weight of a baby because it is known that maternal age influences the birth weight of babies as seen in published articles <br>
<br>

**Backward selection using p-value < 0.10** <br>
```{r}
mod2 <- lm(BWT ~ AGE+LWT+RACE+SMOKE+PTL+HT+UI+FTV, data=df)
summary (mod2)  ## full model
```
<br>
remove variables with p-value > 0.1
<br>
```{r}
clean.mod2 <- update(mod2, . ~ . - FTV, data = df)
summary(clean.mod2)
```
<br>
Age has a p -value greater than 0.1 but we will include it in the model because it is known that maternal age affects birth weight of babies (a condition known as Macrosomia). <br>

The model fitted by backward selection technique using control p-value less than 0.10 is:<br>
**BWT ~ AGE + LWT + RACE + SMOKE + PTL + HT + UI**. This model is similar to the model fitted using forward selection technique and AICc score .
<br>

<br>
**Conclusion**<br>
The final model using AICc criteria is **BWT ~ AGE + LWT + RACE + SMOKE + PTL + HT + UI**. <br>
```{r}
final_model = lm(BWT ~ AGE + LWT + RACE + SMOKE + PTL + HT + UI, data=df)
summary(final_model)
```
<br>
**Confidence interval** <br>
```{r}
cbind(Estimate=final_model$coefficients,confint(final_model))
```
<br>
The coefficients of the predictors are: <br>
**BWT = 2881.17 - 2.29(Age) + 4.38(LWT) - 454.40(RACE2) - 317.19(RACE3) - 331.65(SMOKE1) - 288.59(PTL1) + 1271.59(PTL3) - 576.64(HT1) - 543.57(UT1)**. <br>
The estimated values of the predictors fall within 95% confidence interval of the mean value of the population. Approximately 24 % of the variance in birth weight of newly born babies can be explained by the predictors in the model. <br>

**Interpretation of the model** <br>
* A unit increase in maternal age will likely decrease the birth weight of a newborn baby by 2.29 grams, if all other variables are kept constant <br>
* A unit increase in weight of the woman at her last menstrual period will likely increase the birth weight of a newborn baby by 4.38 grams, if all other variables are kept constant <br>
* The birth weight of  a baby will likely decrease by 454.40 grams if the mother has black ancestry, while keeping all other variables constant. <br>
* The birth weight of  a baby will likely decrease by 317.19 grams if the mother has neither black or white ancestry, while keeping all other variables constant. <br>
* The birth weight of  a baby will likely decrease by 331.65 grams if the mother smoked cigarettes during pregnancy, while keeping all other variables constant. <br>
* The birth weight of  a baby will likely decrease by 228.59 grams if the mother had one history of premature labor term, while keeping all other variables constant. <br>
** The birth weight of  a baby will likely increase by 1271.89 grams if the mother had three history of premature labor term, while keeping all other variables constant. <br>
* The birth weight of  a baby will likely decrease by 576.64 grams if the mother has a history of hypertension, while keeping all other variables constant. <br>
* The birth weight of  a baby will likely decrease by 543.57 grams if the mother has uterine irritability, while keeping all other variables constant. <br>















